Developing a Bubble Chamber Particle Discriminator Using Semi-Supervised Learning

### 2019-02-14 documentation/report.tex
Minor fix
Edits for peer review and stuff; updated research plan

My objective was to develop novel machine learning algorithms that are more accurate, and I succeeded at this goal! I composed a 26-page research whitepaper \cite{me} about my work on PICO-60, which has been approved by the PICO collaboration and pre-published at \url{https://arxiv.org/abs/1811.11308}.

@@

\begin{enumerate}
    \item I experimented further with supervised learning, exploring the following data formats to learn which was the best-performing solution:
    \begin{itemize}
        \item An 8-band Fourier transform of audio recorded in the bubble chamber. I applied a dense neural network with dropout \cite{dropout} and L2 regularization (which allowed for less overfitting to impure data).
        \item A full-resolution Fourier transform of the audio (all 50,001 data points). I once again applied a dense neural network with the Adam \cite{adam} optimizer.
        \item A raw audio waveform with a very deep 1D convolutional neural network (CNN) inspired by Dai et al \cite{verydeepconvnets}.
        \item Images captured by cameras in the detector. I trained a 2D CNN on these, to learn whether they contain any relevant information.
    \end{itemize}
    \begin{figure}[ht]
        \centering
        \subfloat{\fbox{\includegraphics[width=0.9\textwidth]{audio_cropped}}}
        \qquad
        \subfloat{\fbox{\includegraphics[width=0.45\textwidth]{image_grid}}}
        \caption{Examples of the audio waveform and Fourier transform (above) and the images captured by cameras (below).}
    \end{figure}
    \item To even better handle impurities, I developed two novel semi-supervised learning algorithms. The impure alpha/WIMP-like labels were removed from a large portion of the training data, and my algorithms learned more accurate labels for this data.
    \begin{itemize}
        \item Iterative Cluster Nucleation is inspired by unsupervised clustering algorithms. A neural network is trained on a labeled set, and runs inference on the unlabeled set. Unlabeled examples with predictions closest to 0 or 1 are added to the labeled set; the newly synthesized labels are more accurate than the original labels.
        \item Gravitational Differentiation uses a novel piecewise exponential function to calculate final-layer gradients. Unlabeled examples ``gravitate'' toward more accurate predictions based on the confidence of the neural network's predictions.
    \end{itemize}
\end{enumerate}

@@

\begin{itemize}
    \item The dense neural network, with banded Fourier transform data, increased validation accuracy to 97.3\% on validation data, up from the previous ML accuracy of 80.2\%.
    \item The full-resolution Fourier transform unexpectedly worsened accuracy to 92.2\%.
    \item The raw waveform with a 1D very deep CNN reached 94.5\% accuracy.
    \item There was no useful information in the image data; only 63.0\% accuracy was achieved.
    \item Iterative cluster nucleation, using banded FT input data, improved accuracy to 99.1\%.
    \item Gravitational differentiation was even more accurate, at 99.7\%.
    \item The best gravitational differentiation configuration was not as accurate on final test data as on validation data, but was much better than supervised learning, at 98.3\%.
\end{itemize}

@@

\singlespacing
\printbibliography
\doublespacing
\pagebreak

\section*{Bibliography}

>>>>

My objective was to develop novel machine learning algorithms that are more accurate, and I succeeded at this goal! I composed a 28-page research whitepaper \cite{me} about my work on PICO-60, which has been reviewed and approved by the PICO collaboration and pre-published at \url{https://arxiv.org/abs/1811.11308}. It is currently undergoing peer review for eventual publication in the journal \textit{Computer Physics Communications}.

@@

\begin{enumerate}
    \item I experimented further with supervised learning, exploring the following data formats to learn which was the best-performing solution:
    \begin{itemize}
        \item An 8-band Fourier transform of audio recorded in the bubble chamber. I applied a dense neural network with dropout \cite{dropout} and L2 regularization (which allowed for less overfitting to impure data).
        \item A full-resolution Fourier transform of the audio (all 50,001 data points). I once again applied a dense neural network with the Adam \cite{adam} optimizer.
        \item A raw audio waveform with a very deep 1D convolutional neural network (CNN) inspired by Dai et al.\ \cite{verydeepconvnets}.
        \item Images captured by cameras in the detector. I trained a 2D CNN on these, to learn whether they contain any relevant information.
    \end{itemize}
    \begin{figure}[ht]
        \centering
        \subfloat{\fbox{\includegraphics[width=0.9\textwidth]{audio_cropped}}}
        \qquad
        \subfloat{\fbox{\includegraphics[width=0.45\textwidth]{image_grid}}}
        \caption{Examples of the audio waveform and Fourier transform (above) and the images captured by cameras (below).}
    \end{figure}
    \item To even better handle impurities, I developed two novel semi-supervised learning algorithms. The impure alpha/WIMP-like labels were removed from a large portion of the training data, and my algorithms learned more accurate labels for this data.
    \begin{itemize}
        \item Iterative Cluster Nucleation is inspired by unsupervised clustering algorithms. A neural network is trained on a labeled set, and runs inference on the unlabeled set. Unlabeled examples with predictions closest to 0 or 1 are added to the labeled set; the newly synthesized labels are more accurate than the original labels.
        \item Gravitational Differentiation uses a novel piecewise exponential function to calculate final-layer gradients. Unlabeled examples ``gravitate'' toward more accurate predictions based on the confidence of the neural network's predictions.
    \end{itemize}
\end{enumerate}

@@

\begin{itemize}
    \item The dense neural network, with banded Fourier transform data, increased validation accuracy to 97.3\% on validation data, up from the previous ML accuracy of 80.2\%.
    \item The full-resolution Fourier transform produced validation accuracy of 92.2\%.
    \item The raw waveform with a 1D very deep CNN reached 94.5\% accuracy.
    \item There was no useful information in the image data; only 63.0\% accuracy was achieved.
    \item Iterative cluster nucleation, using banded FT input data, improved accuracy to 99.1\%.
    \item Gravitational differentiation was even more accurate, at 99.7\%.
    \item The best gravitational differentiation configuration was not as accurate on final test data as on validation data, but was much better than supervised learning, at 98.3\%.
\end{itemize}

@@

\section{References}

\printbibliography[heading=none]

\section{Bibliography}
